---
title: "1. CensusData_tidycensus"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Working with Census Data in R

## Using the *tidycensus* library

#### Jeremy Mikecz Research Data Services Dartmouth Libraries

In this notebook, we will experiment with importing U.S. Census using the tidycensus package. We will then perform some basic analysis and mapping of a few variables in this dataset.

## Objectives

By the end of the lesson, we will:

-   understand how to access recent data using R, the Census API, and the *tidycensus* library

-   import, wrangle, and analyze / visualize census data using these tools

-   brainstorm a population-related question, find the relevant census data, and answer the question using this data

## Possible Questions

1.  To what extent is New Hampshire's population aging?
2.  ??
3.  ??

## 1. Setup

Things to do in advance of lesson:

1.  install RStudio and R (if not already)

2.  within RStudio, install the packages listed below (in Step 2)

3.  request a census API key (see instructions below)

```{r}

packrat::on()

```

### 1b. TO ACQUIRE A CENSUS API KEY

1.  Request a key [here:](https://api.census.gov/data/key_signup.html)
2.  Save the key in a plain text file that you save as ".Renviron" in the root directory of this project. In the top line of this document, record the API Key like this:

`CENSUS_API_KEY={insert API key here}`

3.  Note: Unlike Python, in R you do not place quotes around the API Key.

<!-- -->

3.  Note #2: The [.Renviron file must end with a new line](https://rstats.wtf/r-startup.html) so press RETURN after the line with the key to ensure there is a blank line at the end of this file. If you're using an editor that shows line numbers, there should be two lines, where the second one is empty.
4.  Restart R. .Renviron is only processed at the start of an R session. As an alternative, you can also run \`readRenviron("\~/.Renviron") to reload your environment without restarting.
5.  After restarting R, open the census key from .Renviron with the commands below:

```{r}
#readRenviron("./.Renviron")
Sys.getenv("CENSUS_API_KEY")  
```

```{r}
getwd()
```

## 2. Datasets

Historical census data may be downloaded from data repositories like **ICPSR** and **IPUMS** for the United States, starting with the first decennial census in 1790. This process typically involves:

-   registering for an account with those repositories

-   downloading census data one census year at a time (and some more recent censuses are divided across multiple datasets).

-   Importing these datasets into R and transforming them into a readable format

Fortunately, if you want data from more recent censuses, you may import data directly into R using the Census API. Even better yet, we can use the **tidycensus** R library to make the import and analysis of this data even easier. [Click here for tidycensus documentation](https://walker-data.com/tidycensus/). Tidycensus is designed for use with the [tidyverse](https://walker-data.com/tidycensus/) packages designed as a more efficient and intuitive way to do data science in R. Hence, we need to install (necessary only once) and load (need to do each time you restart R) these two libraries:

```{r}
#INSTALL PACKAGES - only needed the first time you use this package
#install.packages("tidyverse") 
#install.packages("tidycensus")  

# LOAD THEM
library(tidyverse) 
library(tidycensus) 

```

First, however, let's explore what data is accessible using tidycensus:

1.  U.S. Decennial Census Data - full-count data recorded every 10 years since 1790.

    ```         
    "sf1": 2000 and 2010

    "sf2": 2000 and 2010

    "sf3" and "sf4": 2000 only

    "dp": 2020 only

    "pl" - redistricting files: 2000, 2010, 2020 (for 1990: message: "we recommend using NHGIS and the ipumsr R package.")
    ```

    2.  American Community Survey (ACS) data - survey data recorded more frequently to fill in gaps missed by the decennial census (due to more frequent implementation and wider range of questions asked). Compiled from a survey of 3 million Americans. Because each data point is an estimate, tidycensus returns a margin of error for each variable requested.

```         
5-year ACS ("acs5", default option)
1-year ACS ("acs1")
```

### 2b. Geographic Units / Scales

Through tidycensus, we may request decennial census and ACS data at a variety of geographic levels, from the level of the country, region, and state down to the level of census tracts and blocks. Also includes various alternative geographic categories from Native American reservations to congressional districts, zip code areas, and school districts.

See the [Geography in tidycensus table](https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus) in the package's documentation for a useful overview of these different geographies.
