---
title: "tidyverse intro"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Introduction to Tidyverse

[tidyr cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf)

For a more detailed introduction to working with tidy data in R see:

-   Wickham, Çetinkaya-Rundel, and Grolemund, *R for Data Science*, Ch. 5 ["Data Tidying"](https://r4ds.hadley.nz/data-tidy).

-   Grolemund, [A Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/) (2020).

-   Silge and Robinson, [Text Mining with R](https://www.tidytextmining.com/), Ch. 1 "The tidy text format".

## *Optional*: Working with Virtual Environments

If you are using R for a variety of different projects, you may risk encountering dependency conflicts. These occur when you may need Package *A* for a text analysis project and that package requires version 1.0 of Package *B*, while Package *C* for a machine-learning projects requires version 2.0 of Package *B*.

Setting up a virtual environment allows you to avoid such conflicts. Often, programmers use a different virtual environment for each project (and each project should be stored in a separate project folder). This also provides another benefit. When sharing code with others, programmers can send a list of packages (and their specific versions) to collaborators so that they can reproduce the exact same environment on their machines. This, in turn, ensures that the experience of running the team's code is exactly the same for project members each working on their own computers.

In Python, users typically use the **venv** library to create and use virtual environments. In R, users often use the **Packrat** library / package. To set up a local virtual environment with packrat, you must first install packrat using `install.packages("packrat")`. Then, from within your project folder, initialize it using the command `packrat::init()`. Then, R should activite your new virtual environment automatically whenever you work in that project folder. To ensure it is activated, however, you can run the code `packrat::on()`. To deactivate, simply run `packrat::off()`.

```{r}
#packrat::off()
packrat::on()
#packrat::snapshot()
#packrat::search_path()=
```

```{r}
#tidyverse_update()
library(tidyverse)
#tidyverse_packages()
```

```{r}
starwars

```

## Combine functions into a pipe

Using the symbol `%>%` we can chain multiple functions together. Each new function in the pipeline operates on the results produced by the previous function.

Chaining functions together in a pipe like this:

```         
starwars %>% 
  group_by(species) %>% 
  summarise(avg_height = mean(height, na.rm = TRUE)) %>% 
  arrange(avg_height)
```

produces the same results as nesting a series of functions within another:

```         
arrange(
  summarise(
    group_by(starwars, species), 
    avg_height = mean(height, na.rm = TRUE)
  ), 
  avg_height
)
```

or calling each function in order:

```         
x1 <- starwars 
x2 <- group_by(x1, species)
x3 <- summarise(x3, avg_height = mean(height, na.rm = TRUE))
arrange(x3, avg_height)
```

```{r}
starwars %>%
  group_by(species)
```

```{r}
starwars %>%
  group_by(species) %>%
  summarise(avg_height = mean(height, na.rm = TRUE)) 
```

```{r}
starwars %>% 
  group_by(species) %>% 
  summarise(avg_height = mean(height, na.rm = TRUE)) %>% 
  arrange(avg_height)

```

By default `%>%` passes the result of the left hand side to the the first unnamed argument of the function on the right hand side. To override this default, use `.` as a placeholder within the function call on the right hand side. `%>%` will evaluate `.` as the result of the left hand side, instead of passing the result to the first unnamed argument.

```{r}
starwars %>%
  lm(mass ~ height, data = .)
```

```{r}
data <- read_csv("census/data/census1970.csv")
data
```

```{r}
df <- read.csv("census/data/census1970.csv")
df

```

## Tidy Data

From the [Tidyverse cookbook](https://rstudio-education.github.io/tidyverse-cookbook/tidy.html):

Data tidying refers to reshaping your data into a tidy data frame or [tibble](https://rstudio-education.github.io/tidyverse-cookbook/how-to-use-this-book.html#tidy-data). Data tidying is an important first step for your analysis because every tidyverse function will expect your data to be stored as **Tidy Data**.

Tidy data is tabular data organized so that:

1.  Each column contains a single variable

2.  Each row contains a single observation

Tidy data is not an arbitrary requirement of the tidyverse; it is the ideal data format for doing data science with R. Tidy data makes it easy to extract every value of a variable to build a plot or to compute a summary statistic. Tidy data also makes it easy to compute new variables; when your data is tidy, you can rely on R's rowwise operations to maintain the integrity of your observations. Moreover, R can directly manipulate tidy data with R's fast, built-in vectorised observations, which lets your code run as fast as possible.

The definition of Tidy Data isn't complete until you define variable and observation, so let's borrow two definitions from [*R for Data Science*](https://r4ds.had.co.nz/exploratory-data-analysis.html):

1.  A **variable** is a quantity, quality, or property that you can measure.

2.  An **observation** is a set of measurements made under similar conditions (you usually make all of the measurements in an observation at the same time and on the same object).

As you work with data, you will be surprised to realize that what is a variable (or observation) will depend less on the data itself and more on what you are trying to do with it. With enough mental flexibility, you can consider anything to be a variable. However, some variables will be more useful than others for any specific task. In general, if you can formulate your task as an equation (math or code that contains an equals sign), the most useful variables will be the names in the equation.

### Create a tibble from scratch

```{r}
omitb <- tribble(~name, ~occupation, ~apartment, ~is_suspect, ~is_dead,
            "Charles", "washed-up actor", "12D", "no", "no",
            "Mabel", "unemployed never-war", "14C", "no", "no",
            "Oliver", "theater director", "10A", "no", "no",
            "Howard", "childless cat guy", "6F", "yes", "no",
            "Bunny", "petty despot", "1A", "no", "maybe"
)
omitb
```

Use the **tribble** format to create a tibble by using a syntax that lines up the rows neatly in a way that will preview the result.

You can also create the same tibble using the **tibble** function this way

```{r}
tibble(name = c("Charles", "Mabel", "Oliver", "Howard", "Bunny"),
       occupation = c("washed-up actor", "unemployed never-was", "theater director", "childless cat guy", "petty despot"),
       apartment = c("12D", "14C", "10A", "6F", "1A"),
       is_suspect = c("no", "no", "no", "yes", "no"),
       is_dead = c("no", "no", "no", "no", "maybe"),
       )
```

## Dataframes vs. tibbles

### Convert a dataframe into a tibble

```{r}
head(df)

```

```{r}
df2tib <- as_tibble(df)
```

## Sort Table (tibble)

```{r}
starwars %>%
  arrange(homeworld)  #descending order: arrange(desc(homeworld))
```

## Filter Table

```{r}
starwars %>%
  filter(species=="Droid")
```

## Subset Tibble by Column Names

```{r}
starwars %>%
  select(name, homeworld, species)  #reorder names here to reorder columns

```

### For a range of columns

```{r}
starwars %>% 
  select(name:hair_color, species)

```

### Rename columns

```{r}
starwars %>%
  rename(character = name, planet = homeworld)
```

```{r}
starwars$vehicles[1]
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
