---
title: "NH_CensusData"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Working with Census Data in R

In this notebook, we will experiment with importing U.S. Census using the tidycensus package. We will then perform some basic analysis and mapping of a few variables in this dataset.

Things to do in advance of lesson:

1.  install RStudio and R (if not already)

2.  within RStudio, install the packages listed below

3.  request a census API key (see instructions below)

```{r}
#install.packages("tidycensus")
#library(tidyr)
library(tidyverse)
library(tidycensus)

```

## TO ACQUIRE A CENSUS API KEY

1.  Request a key [here:](https://api.census.gov/data/key_signup.html)
2.  Save the key in a plain text file that you save as ".Renviron" in the root directory of this project. In the top line of this document, record the API Key like this:

`CENSUS_API_KEY={insert API key here}`

3.  Note: Unlike Python, in R you do not place quotes around the API Key.

<!-- -->

3.  Note #2: The [.Renviron file must end with a new line](https://rstats.wtf/r-startup.html) so press RETURN after the line with the key to ensure there is a blank line at the end of this file. If you're using an editor that shows line numbers, there should be two lines, where the second one is empty.
4.  Restart R. .Renviron is only processed at the start of an R session. As an alternative, you can also run \`readRenviron("\~/.Renviron") to reload your environment without restarting.
5.  After restarting R, open the census key from .Renviron with the commands below:

```{r}
Sys.getenv("CENSUS_API_KEY")


```

## Datasets:

### Decennial Census datasets

"sf1": 2000 and 2010

"sf2": 2000 and 2010

"sf3" and "sf4": 2000 only

"dp": 2020 only

"pl" - redistricting files: 2000, 2010, 2020 (for 1990: message: "we recommend using NHGIS and the ipumsr R package.")

```{r}

NH_congdists <- get_decennial(geography = "congressional district", 
                         state = "NH", 
                         variables = c("P5_002N"),
                         year = 2020)
head(NH_congdists)
```

### American Community Survey (ACS) datasets

According to the [tidycensus documentation](https://walker-data.com/tidycensus/articles/basic-usage.html) ACS 1-year data is available since 2005 for geographic units with a population of 65,000 or greater and from the 5-year ACS for all geographies down to the block group level, starting with the 2005-2009 period.

```{r}
#vars <- load_variables(2020, "pl", cache = TRUE)
vars <- load_variables(2022, "acs5", cache = TRUE)
view(vars)

write.csv(vars, "acs5_2022_variables.csv", row.names=FALSE)

```

```{r}
?load_variables

```

get_acs variables do not include the "congressional district" geography, even though \[tidycensus documentation\](https://walker-data.com/tidycensus/articles/basic-usage.html) says it should.

```{r}
vars %>%
  distinct(geography)


```

[overview of U.S. census geography](https://learn.arcgis.com/en/related-concepts/united-states-census-geography.htm)

```{r}
get_acs(geography = "tract", 
                         state = "NH", 
                         variables = c("B01001A_001"),
                         year = 2022)

```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```
