---
title: "R Notebook"
output: html_notebook
---



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
##WORKING WITH API
# https://www.hrecht.com/censusapi/articles/example-list.html#american-community-survey
# https://learn.r-journalism.com/en/mapping/census_maps/census-maps/
# https://rconsortium.github.io/censusguide/r-packages-all.html
setwd("C:/Users/F0040RP/documents/DartLib_RDS/censusData")
# install.packages("tidycensus")
library(tidycensus)
library(tidyverse)
# install.packages("readr")
library(readr)
# install.packages("censusapi")
library(censusapi)
library(dplyr)
# install.packages("tigris")
library(tigris)
# install.packages("leaflet")
library(leaflet)
#To enable caching of data, set `options(tigris_use_cache = TRUE)`
# in your R script or .Rprofile.
fileName<-"census_api.txt"
# c<-readChar(fileName,file.info(fileName))
c<-read_file(fileName)


# census_api_key(c, install = TRUE, overwrite=TRUE)  #install - only have to run this line once
#problem census api uses CENSUS_KEY instead of CENSUS_API_KEY
Sys.setenv(CENSUS_KEY=c)
# census_key(c,install=TRUE)
# First time, reload your environment so you can use the key without restarting R.
# readRenviron("~/.Renviron")
# You can check it with:
# Sys.getenv("CENSUS_API_KEY")

Sys.getenv("CENSUS_KEY")



```


Explore datasets available
```{r}
apis<-listCensusApis() #only goes back to 80s? decenniel census only back to 2000!!; 

data2020 <- getCensus(
    name = "dec/pl",
    vintage = 2020,
    vars = c("NAME", "P1_001N"), 
    region = "place:*",
    regionin = "state:05")

head(data2020)
```

Get total population and the number of housing units for metropolitan/micropolitan statistical areas in 2010.
```{r}
data2010 <- getCensus(
    name = "dec/sf1",
    vintage = 2010,
    vars = c("NAME", "P001001", "H010001"), 
    region = "metropolitan statistical area/micropolitan statistical area:*")

colnames(data2010) <- c("MSA", "NAME", "population","housingUnits")
head(data2010)

```
# Show variable metadata for the P2 group
```{r}
group_p2 <- listCensusMetadata(
    name = "dec/sf1",
    vintage = 2010,
    type = "variables",
    group = "P2")
```


# Get the P2 variable group (URBAN AND RURAL)
```{r}
data2010b <- getCensus(
    name = "dec/sf1",
    vintage = 2010,
    vars = "group(P2)", 
    region = "metropolitan statistical area/micropolitan statistical area:*")
head(data2010b)

```

```{r}
tract_pop <- getCensus(
    name = "dec/sf1",
    vintage = 2010,
    vars = "P001001", 
    region = "block:*",
    regionin = "state:36+county:027+tract:010000")
head(tract_pop)


```


```{r}

acs_income <- getCensus(
    name = "acs/acs5",
    vintage = 2020, 
    vars = c("NAME", "B01001_001E", "B19013_001E"), 
    region = "place:*",
    regionin = "state:18")
head(acs_income)

```

Get one variable from 1 ACS year
```{r}
state_pop <-  getCensus(name="acs/acs5", 
                        vintage=2015,
                        key=c, 
                        vars=c("NAME", "B01003_001E"), 
                        region="state:*")

head(state_pop[order(-state_pop$B01003_001E),])
```

Cleaning up the column names

```{r}
colnames(state_pop) <- c("state_id", "NAME", "population")
state_pop$state_id <- as.numeric(state_pop$state_id)
state_pop<-state_pop[order(-state_pop$population),]
head(state_pop)

```
# Hm, data comes in numbers of fully spelled out, not abbreviations

# Did you know R has its own built in list of State names and State abbreviations?
# Just pull it in this way to create a dataframe for reference

```{r}
state_off <- data.frame(state.abb, state.name)
head(state_off)
```


