---
title: "R Notebook"
output: html_notebook
---



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
##WORKING WITH API
#https://walker-data.com/tidycensus/articles/basic-usage.html
# https://www.ncdemography.org/2022/05/16/story-recipe-how-to-obtain-census-data-using-r-tidycensus/
#R packages that work with census data: https://rconsortium.github.io/censusguide/r-packages-all.html
setwd("C:/Users/F0040RP/documents/DartLib_RDS/censusData")
# install.packages("tidycensus")
library(tidycensus)
library(tidyverse)
# install.packages("readr")
library(readr)
fileName<-"census_api.txt"
# c<-readChar(fileName,file.info(fileName))
c<-read_file(fileName)


# census_api_key(c, install = TRUE, overwrite=TRUE)  #install - only have to run this line once
# First time, reload your environment so you can use the key without restarting R.
readRenviron("~/.Renviron")
# You can check it with:
Sys.getenv("CENSUS_API_KEY")

```

get one variable across states

```{r}
age10 <- get_decennial(geography = "state", 
                       variables = "P013001", 
                       year = 2010)
head(age10)

```

graph average age by state

```{r}
age10 %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()

```
graph average age by county
```{r}
age10c <- get_decennial(geography = "county", 
                       variables = "P013001", 
                       year = 2010)

age10c %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()


```
average age by census tract
```{r}
age10t_nh <- get_decennial(geography = "tract", 
                        state="NH",
                        variables = "P013001", 
                        year = 2010)

age10t_nh %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()

```

Load variables

```{r}
decennial_2020_vars <- load_variables(
  year = 2020, 
  "pl", 
  cache = TRUE
)

# 2010 Decennial Census Variables
decennial_2010_vars <- load_variables(
  year = 2010, 
  "pl", 
  cache = TRUE
)


# varlist <- load_variables(
#   year = 1980, 
#   "pl", 
#   cache = TRUE
# )

#load variables -ACS data:
v17 <- load_variables(2017, "acs5", cache = TRUE)

# 2016 - 2020 5 Year American Community Survey (ACS) Variables
acs_20_vars = load_variables(
  year = 2020, 
  "acs5",
  cache = TRUE
)

```

CREATE A LIST OF DESIRED VARIABLES:
```{r}
desired_vars = c(
  all = "P2_001N",
  hisp = "P2_002N",
  white = "P2_005N",
  baa = "P2_006N",
  amin = "P2_007N",
  asian = "P2_008N",
  nhopi = "P2_009N",
  other = "P2_010N",
  multi = "P2_011N"
)

```
retrieve census data
```{r}
census_data = get_decennial(
  geography = "county",
  state = "NC",
  variables = desired_vars, #<-- here is where I am using the list
  summary_var = "P2_001N", # <-- creates a column w/'total' variable
  year = 2020,
  sumfile = "pl"
)

```

retrieve ACA data
```{r}
nc_county_income = get_acs(
  geography = "county",
  state = "NC",
  table = "B19001")## Note that leaving the 'year' argument blank tells the API 
  #to return the most recent year available. As of writing this, that is 2020 
  #for both the ACS a

nh_medinc <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "NH", 
              year = 2018)

```
graphing margin of error:
```{r}
nh_medinc %>%
  mutate(NAME = gsub(" County, New Hampshire", "", NAME)) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household income by county in New Hampshire",
       subtitle = "2014-2018 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```


```{r}

```


```{r}

```


```{r}

```


```{r}

```