---
title: "R Notebook"
output: html_notebook
---



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
##WORKING WITH API
# https://learn.r-journalism.com/en/mapping/census_maps/census-maps/
# https://rconsortium.github.io/censusguide/r-packages-all.html
setwd("C:/Users/F0040RP/documents/DartLib_RDS/censusData")
# install.packages("tidycensus")
library(tidycensus)
library(tidyverse)
# install.packages("readr")
library(readr)
# install.packages("censusapi")
library(censusapi)
library(dplyr)
# install.packages("tigris")
library(tigris)
# install.packages("leaflet")
library(leaflet)
#To enable caching of data, set `options(tigris_use_cache = TRUE)`
# in your R script or .Rprofile.
fileName<-"census_api.txt"
# c<-readChar(fileName,file.info(fileName))
c<-read_file(fileName)


# census_api_key(c, install = TRUE, overwrite=TRUE)  #install - only have to run this line once
# First time, reload your environment so you can use the key without restarting R.
# readRenviron("~/.Renviron")
# You can check it with:
Sys.getenv("CENSUS_API_KEY")

```


Get one variable from 1 ACS year
```{r}
state_pop <-  getCensus(name="acs/acs5", 
                        vintage=2015,
                        key=c, 
                        vars=c("NAME", "B01003_001E"), 
                        region="state:*")

head(state_pop[order(-state_pop$B01003_001E),])
```

Cleaning up the column names

```{r}
colnames(state_pop) <- c("state_id", "NAME", "population")
state_pop$state_id <- as.numeric(state_pop$state_id)
state_pop<-state_pop[order(-state_pop$population),]
head(state_pop)

```
# Hm, data comes in numbers of fully spelled out, not abbreviations

# Did you know R has its own built in list of State names and State abbreviations?
# Just pull it in this way to create a dataframe for reference

```{r}
state_off <- data.frame(state.abb, state.name)
head(state_off)
```

Bringing in and Mapping Other Sample Data
```{r}
# install.packages("usethis")
# usethis::use_course("https://github.com/r-journalism/learn-chapter-2/archive/master.zip")
# usethis::use_course("https://github.com/r-journalism/learn-chapter-5/archive/master.zip")

setwd("C:/Users/F0040RP/Documents/DartLib_RDS/censusData/learn-chapter-5-master/census_maps")

dunkin <- read_csv("data/dunkin.csv")
starbucks <- read.csv("data/starbucks.csv", stringsAsFactors=F)




```
https://learn.r-journalism.com/en/mapping/leaflet_maps/leaflet/

```{r}

# Insert your latitude and longitude in the code below
# NOTE: Don't get them reversed otherwise you'll end up in the South Pole.

# Initialize and assign m as the leaflet object
m <- leaflet() %>%
# Now add tiles to it
    addTiles() %>%  
# Setting the middle of where the map should be and the zoom level
    setView(lng=-77.030137, lat=38.902986, zoom = 16) %>%
# Now, add a marker with a popup, 
    addMarkers(lng=-77.030137, lat=38.902986, popup="<b>Hello</b><br><a href='https://www.washingtonpost.com'>-Me</a>")

m 

```

```{r}
glimpse(dunkin)


```
```{r}
dd_state <- dunkin %>% 
  filter(state=="NH")

m <- leaflet(dd_state) %>% addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  setView(-71.931180, 42.385453, zoom = 7) %>% 
  addCircles(~lon, ~lat, popup=dunkin$type, weight = 3, radius=40, 
                 color="#ffa500", stroke = TRUE, fillOpacity = 0.8) 
m
```

```{r}
sb_state <- starbucks %>% 
  filter(Province=="NH")

# isolating just the 3 columns we're interested in-- type, lat, and lon
sb_loc <- select(sb_state, type, lat, lon)
dd_loc <- select(dd_state, type, lat, lon)

# joining the two data frames together
ddsb <- rbind(sb_loc, dd_loc)

# creating a coffee color palette

cof <- colorFactor(c("#ffa500", "#13ED3F"), domain=c("Dunkin Donuts", "Starbucks"))
# mapping based on type
m <- leaflet(ddsb) %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  setView(-71.931180, 42.385453, zoom = 7) %>% 
  addCircleMarkers(~lon, ~lat, popup=ddsb$type, weight = 3, radius=4, 
                 color=~cof(type), stroke = F, fillOpacity = 0.5) 
m
```
add a legend

```{r}
m <- leaflet(ddsb)  %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  setView(-71.931180, 42.385453, zoom = 7) %>% 
  addCircleMarkers(~lon, ~lat, popup=ddsb$type, weight = 3, radius=4, 
                 color=~cof(type), stroke = F, fillOpacity = 0.5)  %>%
  addLegend("bottomright", colors= c("#ffa500", "#13ED3F"), labels=c("Dunkin'", "Starbucks"), title="Coffee places") 

m
```


MAPPING STATES BY POPULATION

```{r}

sb_state <- starbucks %>%
  group_by(Province) %>%
  summarize(total=n()) %>% 
# Some quick adjustments to the the dataframe to clean up names
  mutate(type = "Starbucks") %>%  
  rename(state=Province)

dk_state <- dunkin %>%
  group_by(state) %>%
  summarize(total=n())

states <- states(cb=T)
states %>% 
  leaflet() %>% 
  addTiles() %>% 
  addPolygons(popup=~NAME)

# Now we use the Tigris function geo_join to bring together 
# the states shapefile and the sb_states dataframe -- STUSPS and state 
# are the two columns they'll be joined by

states_merged_sb <- geo_join(states, sb_state, "STUSPS", "state")

# Creating a color palette based on the number range in the total column
pal <- colorNumeric("Greens", domain=states_merged_sb$total)

# Getting rid of rows with NA values
# Using the Base R method of filtering subset() because we're dealing with a SpatialPolygonsDataFrame and not a normal data frame, thus filter() wouldn't work

states_merged_sb <- subset(states_merged_sb, !is.na(total))

# Setting up the pop up text
popup_sb <- paste0("Total: ", as.character(states_merged_sb$total))

# Mapping it with the new tiles CartoDB.Positron
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-98.483330, 38.712046, zoom = 4) %>% 
  addPolygons(data = states_merged_sb , 
              fillColor = ~pal(states_merged_sb$total), 
              fillOpacity = 0.7, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = ~popup_sb) %>%
  addLegend(pal = pal, 
            values = states_merged_sb$total, 
            position = "bottomright", 
            title = "Starbucks")
```



Hmm… Not that interesting, right?

What’s the problem here. Deep down, you know what’s wrong.

This is essentially a population map.

So we need to adjust for population.

And that’s easy to do using the Census API.
```{r}
# So I needed to create the dataframe above because the Census API data 
# gave me states with full names while the Starbucks data came with abbreviated state names
# So I needed a relationship dataframe so I could join the two

# Cleaning up the names for easier joining
colnames(state_off) <- c("state", "NAME")

# Joining state population dataframe to relationship file
state_pop <- left_join(state_pop, state_off)

# The relationship dataframe didnt have DC or Puerto Rico, so I'm manually putting those in
state_pop$state <- ifelse(state_pop$NAME=="District of Columbia", "DC", as.character(state_pop$state))
state_pop$state <- ifelse(state_pop$NAME=="Puerto Rico", "PR", as.character(state_pop$state))

# Joining Starbucks dataframe to adjusted state population dataframe
sb_state_pop <- left_join(sb_state, state_pop)

# Calculating per Starbucks stores 100,000 residents and rounding to 2 digits
sb_state_pop$per_capita <- round(sb_state_pop$total/sb_state_pop$population*100000,2)

# Eliminating rows with NA
sb_state_pop <- filter(sb_state_pop, !is.na(per_capita))
head(sb_state_pop[order(-sb_state_pop$per_capita),])
```

```{r}
states_merged_sb_pc <- geo_join(states, sb_state_pop, "STUSPS", "state")

pal_sb <- colorNumeric("Greens", domain=states_merged_sb_pc$per_capita)
states_merged_sb_pc <- subset(states_merged_sb_pc, !is.na(per_capita))

# Here's the pop up
popup_sb <- paste0("<strong>", states_merged_sb_pc$NAME, 
                   "</strong><br />Total: ", states_merged_sb_pc$total,
                   "<br />Per capita: ", 
                   as.character(states_merged_sb_pc$per_capita))
head(popup_sb)

```



```{r}
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(-98.483330, 38.712046, zoom = 4) %>% 
  addPolygons(data = states_merged_sb_pc , 
              fillColor = ~pal_sb(states_merged_sb_pc$per_capita), 
              fillOpacity = 0.9, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = ~popup_sb) %>%
  addLegend(pal = pal_sb, 
            values = states_merged_sb_pc$per_capita, 
            position = "bottomright", 
            title = "Starbucks<br />per 100,000<br/>residents")
```

Now do with Dunkin
```{r}
# Joining Starbucks dataframe to adjusted state population dataframe
dk_state_pop <- left_join(dk_state, state_pop)

# Calculating per Starbucks stores 100,000 residents and rounding to 2 digits
dk_state_pop$per_capita <- round(dk_state_pop$total/dk_state_pop$population*100000,2)

# Eliminating rows with NA
dk_state_pop <- filter(dk_state_pop, !is.na(per_capita))
head(dk_state_pop[order(-dk_state_pop$per_capita),])
```




```{r}
states_merged_dk_pc <- geo_join(states, dk_state_pop, "STUSPS", "state")

pal_dk <- colorNumeric("Greens", domain=states_merged_dk_pc$per_capita)
states_merged_dk_pc <- subset(states_merged_dk_pc, !is.na(per_capita))

# Here's the pop up
popup_dk <- paste0("<strong>", states_merged_dk_pc$NAME, 
                   "</strong><br />Total: ", states_merged_dk_pc$total,
                   "<br />Per capita: ", 
                   as.character(states_merged_dk_pc$per_capita))
head(popup_dk)
```



```{r}
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(-98.483330, 38.712046, zoom = 4) %>% 
  addPolygons(data = states_merged_dk_pc , 
              fillColor = ~pal_dk(states_merged_dk_pc$per_capita), 
              fillOpacity = 0.9, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = ~popup_dk) %>%
  addLegend(pal = pal_dk, 
            values = states_merged_dk_pc$per_capita, 
            position = "bottomright", 
            title = "Starbucks<br />per 100,000<br/>residents")
```



```{r}

```
