---
title: "R Basics"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R Basics for the Beginning R Programmer

## Basic Commands
```{r}
print("Hello World!")
```

## Basic Math

Analyze the following operands. Then describe what they do in a comment (starting with "#") next to each line of code.
```{r}
8*3.33  
67/5    
246+632
1156-1256 

26/4      
26 %/%4  #%/% = 
26 %% 4  #%% = 

4/2==2   #== (equals)
3+7==8   

3^2      #^ = 
3^3
3^4
sqrt(9)  #sqrt = 

```
## Storing to Variables

Note: in many programming languages, you would assign a variable using the equal ("=") sign.

You may also do that in R, but best practices suggest using "<-" instead. Do an online search for "R = vs <-" and you'll see a lot of discussion about this. 

```{r}
x<- 6
y<- 3 
#one problem with the "<-" approach is that there is a key difference betweem the above code (which assigns 3 to the variable y, and the below code, which asks if y is less than negative 3)
y< -3
x*y
x/y
x+y
x^y

#meanwhile:
z<-5.5 #is equivalent to z<- 5 NOT z< -5, but adding the space makes this clear
print
a< -8.3  #this one produces an error because we haven't yet assigned a value to a
b<- 8.3 #but this one doesn't
print(b)
```
We can use more descriptive variables as well:

```{r}
Item<-"thingamajig"
Age<-44
kg<-2.6
heightCM<-68
widthCM<-52
lengthCM<-114

Volm2<-(heightCM/100)*(widthCM/100)*(lengthCM/100)
Volcm2<-Volm2*(100^3)
print(Volm2)
print(Volcm2)
#density (kg/m3) = mass (kg) / volume (m3)
dens<- kg / Volm2
#FIX THIS!! how to print multiple items in R!!
print(c("The thingamajig is",widthCM,"cm x",heightCM,"cm x",lengthCM, "and weighs",kg,"kg."))
print(c("and thus has a density of",dens,"cubic meters (m3)."))
```

Is the area of a circular farm plot, with a radius of 200 meters, greater than that of a square plot with a width of 400 meters?
```{r}
print(pi)
# you may recall from high school geometry that the area of a circle is equal to pi * radius squared (or A=pi*r2)
ciArea = pi*200^2
sqArea = 400^2
ciArea>sqArea
print(c("A circle with a radius of 200m has an area of ",ciArea,"meters squared, whereas a square with the same diameter/width has an area of ",sqArea,"meters squared."))
```

  





## Working with Lists

```{r}

```

## Working with Dataframes
```{r}

```

```{r}

```
