---
title: "Congressional Districts"
format: html
editor: visual
---

## Congressional District Data

## Notes - from previous email

I believe the Census API  allows retrieval of demographic info by congressional district for the years 2000, 2010, and 2020. The ACS API has a lot more variables to look at (but obviously its an estimate rather than a full count).

What types of breakdowns would you want besides 18 and up? 

Race? Sex? Others? race x age?  race x sex ? etc.

I don't believe either the Census or ACS has numbers for registered voters. But, there may be other ways to get this, possibly through the Current Population Survey: <https://www.census.gov/programs-surveys/cps.html> (just found this, haven't had time to explore it yet).

```{r}
packrat::on()

readRenviron("../.Renviron")
Sys.getenv("CENSUS_API_KEY")  

```

```{r}
#INSTALL PACKAGES - only needed the first time you use this package
#install.packages("tidyverse") 
#install.packages("tidycensus")  

# LOAD THEM
library(tidyverse) 
library(tidycensus) 

```

```{r}
vars2020 <- load_variables(2020, "pl", cache=TRUE)
vars2020dhc <- load_variables(2020, "dhc", cache=TRUE)

#2022 - most recent acs5 dataset
vars2022acs <- load_variables(2022, "acs5", cache=TRUE)

#view(vars2020)
```

```{r}
NH_dhc_cds_total18 <- get_decennial(geography = "congressional district",
                             state="NH",
                             sumfile="dhc",
                             variables = c(total18="P10_001N",
                                           race1_18 = "P10_002N",
                                           white1_18 = "P10_003N",
                                           black1_18 = "P10_004N",
                                           race2_18 = "P10_010N",
                                           hisp_18 = "P11_002N",
                                           nonhisp_18="P11_003N"
                                           ),
                             year = 2020)

```

## Black / African-American in Combination with other races

P12AD

```{r}
NH_dhc_cds_blackhisp_plus18 <- get_decennial(geography = "congressional district",
                             state="NH",
                             sumfile="dhc",
                             table = "P12AD",
                             year = 2020)
```

```{r}
#extract last 3 numbers of variable and convert to integer
NH_dhc_cds_blackhisp_plus18 <-NH_dhc_cds_blackhisp_plus18 %>%
  mutate(groupnum = as.numeric(str_sub(variable, -4, -2)))
```

```{r}

#group by over 18

NH_dhc_cds_blackhisp_plus18 <- NH_dhc_cds_blackhisp_plus18 %>%
  mutate(variable=case_when((between(groupnum, 7,25)) | (between(groupnum, 31, 49))
                                                         ~ "blackplus18"))

#NH_dhc_cds_blackplus18 <- NH_dhc_cds_blackplus18 %>%
#  mutate(over18 = case_when((between(groupnum, 7,25)) | (between(groupnum, 31, 49)) ~ TRUE, TRUE ~ FALSE))
#

```

```{r}
NH_dhc_cds_blacknh_plus18 <- get_decennial(geography = "congressional district",
                             state="NH",
                             sumfile="dhc",
                             table = "P12X",
                             year = 2020)

```

```{r}
#extract last 3 numbers of variable and convert to integer
NH_dhc_cds_blacknh_plus18 <-NH_dhc_cds_blacknh_plus18 %>%
  mutate(groupnum = as.numeric(str_sub(variable, -4, -2)))

NH_dhc_cds_blacknh_plus18 <- NH_dhc_cds_blacknh_plus18 %>%
  mutate(variable=case_when((between(groupnum, 7,25)) | (between(groupnum, 31, 49))
                                                         ~ "blackplus18"))

```

```{r}
NH_dhc_cds_blackplus18  <- NH_dhc_cds_blackhisp_plus18 %>%
  bind_rows(NH_dhc_cds_blacknh_plus18)

```

```{r}
#sum up all over 18
NH_dhc_cds_blackplus18group <- NH_dhc_cds_blackplus18 %>% 
  group_by(GEOID, variable, NAME) %>% 
  summarise(value = sum(value)) %>%
  filter(variable!="")
```

## Sex 18+

P12

```{r}
NH_dhc_cds_sex18 <- get_decennial(geography = "congressional district",
                             state="NH",
                             sumfile="dhc",
                             table = "P12",
                             year = 2020)



```

```{r}
  #extract last 3 numbers of variable and convert to integer
NH_dhc_cds_sex18 <-NH_dhc_cds_sex18 %>%
  mutate(groupnum = as.numeric(str_sub(variable, -4, -2)))

#group by over 18
NH_dhc_cds_sex18 <- NH_dhc_cds_sex18 %>%
  mutate(variable = case_when(between(groupnum, 7,25) ~ "male18",
                           between(groupnum, 31, 49) ~ "female18"))

#sum up all over 18
NH_dhc_cds_sex18group <- NH_dhc_cds_sex18 %>% 
  group_by(GEOID, variable, NAME) %>% 
  summarise(value = sum(value)) %>%
  filter(variable!="")
```

## Combine Data

```{r}
'
test <- NH_dhc_cds_total18 %>%
  full_join(NH_dhc_cds_sex18group, by=c("NAME", "variable"))
'


```

```{r}
educ18 <- read.csv("NH_acs_cds_educ18.csv", header=TRUE)
educ18$GEOID <- as.character(educ18$GEOID)
#glimpse(NH_dhc_cds_total18)

```

```{r}
NH_dhc_cds <- NH_dhc_cds_total18 %>%
  bind_rows(NH_dhc_cds_sex18group) %>%
  bind_rows(NH_dhc_cds_blackplus18group) %>%
  bind_rows(educ18) %>%
  select(!X)

write.csv(NH_dhc_cds, "NHcongdistdata_long.csv")

```

```{r}
NH_dhc_cds_wide <- NH_dhc_cds %>%
  mutate(value = case_when(
    is.na(value) ~ estimate, TRUE ~ value))%>%
  select(!estimate) %>%
  pivot_wider(names_from=variable, values_from = c("value", "moe"))



```

```{r}
NHall_dhc <- NH_dhc_cds %>%
  group_by(variable) %>%
  summarise(value = sum(value), estimate = sum(estimate), moe_sum = moe_sum(moe, estimate))

```
